An excess of positive results: Comparing the standard Psychology literature with Registered Reports

### 2020-01-14 preprint/manuscript.Rmd
changed one sentence in the Discussion and two in the description of replication status coding

Assuming a constant alpha level, the rate of positive results in a literature (i.e., the proportion of supported hypotheses among all tested hypotheses) is influenced by three factors: the proportion of true hypotheses among all tested hypotheses, statistical power, and bias. 
The Registered Reports format combines powerful safeguards against publication bias and QRPs with standards for research quality that are at least equal to ordinary peer review, and often include statistical power requirements that likely exceed those in the standard literature [see e.g., @Maxwell2004; @Szucs2017]. 
Therefore, if the emerging Registered Reports literature in Psychology contains fewer positive results than the standard literature, the cause must be either the difference in bias or a lower proportion of true hypotheses tested in Registered Reports (or a combination of the two). 
At this time, we have good reasons to believe in a difference in bias, but less reason to believe in a difference in the proportion of true hypotheses (at least regarding original hypotheses, see below), which would make bias a more plausible explanation of a potential difference in the positive result rate. 
Considering the high standards for research quality in Registered Reports, a large difference in positive results between Registered Reports and the standard literature would also mean that publication bias is not a desirable filter for poorly conducted studies, but that we should worry about the high-quality negative results we are missing because of it.

@@

For standard reports we downloaded a current version of the Essential Science Indicators (ESI) database (retrieved on 4th December 2018) and used Web of Science to search for articles published between 2013 and 2018 with a Boolean search query containing the phrase "test$^\ast$ the hypothes$^\ast$" and the ISSNs of all journals listed in the ESI Psychiatry/Psychology category. 
Using the same sample size as @Fanelli2010, we randomly selected 150 papers from the 1919 search results using the sample() function in R and the seed "20190120" (not specified in our preregistration, but no other seeds were tried).
We initially excluded eight papers and replaced them (the decision to replace excluded papers was not preregistered) through the same random sampling procedure until 150 studies were found that met our criteria, but later found that two papers had erroneously been excluded, leading to a final sample size of 152 rather than 150 (see Fig.\ \@ref(fig:sampling)).

@@

For Registered Reports we aimed to include all published Registered Reports in the field of Psychology that tested at least one hypothesis, regardless of whether or not they used the phrase "test$^\ast$ the hypothes$^\ast$". 
We downloaded a database of published Registered Reports curated by the Center for Open Science (https://www.zotero.org/groups/479248/osf/items/collectionKey/KEJP68G9, retrieved on 19th November 2018), and excluded papers published in journals that were listed in categories other than "Psychiatry/Psychology" or "Multidisciplinary" in the ESI. 
Note that the decision to focus only on the Psychiatry/Psychology category meant excluding 13 Registered Reports published in Cortex because the ESI counts this journal towards the separate category "Neuroscience and Behavior". 
Papers published in multidisciplinary journals and in journals not included in the ESI (e.g., Royal Society Open Science) were hand-coded by AS. 
This deviates from our preregistration insofar as we had not specified how discipline membership would be determined.

@@

### Replication status
We also wanted to code if a study was a replication of a previously published one: 
We expected a much larger proportion of Registered Reports to be direct replications, many of which may have been motivated by scepticism of the original study. 
A lower positive result rate in Registered Reports could then be an effect of failed replications rather than an effect of safeguards against QRPs and publication bias. 
After an initial coding attempt with ill-defined coding criteria had led to too many disagreements (described further in the Appendix), we developed the following strategy (not pre-registered): 
We searched the full texts of all papers for the string "replic$^\ast$" [cf. @Makel2012; @Kohler2019; @Mueller-Langer2019; @Pridemore2018], and for papers that did contain it determined whether the first hypothesis was a close or direct replication of previously published work. 
Conceptual replications and internal replications were not counted as replications in this narrow sense: 
The reason to code replication status was the idea that direct replications of previous work may be motivated by replicators' scepticism in the hypothesis and thus have a low prior probability, whereas conceptual replications are usually considered as building on previous work rather than verifying it. 
AS coded all papers, DL double-coded `r rep.RR.coded.DL` Registered Reports ($`r printnum(rep.RR.coded.DL/sum(included$is_RR)*100)` \%$) and `r rep.SR.coded.DL` standard reports ($`r printnum(rep.SR.coded.DL/sum(included$is_RR==0)*100)` \%$). 
There were `r rep.disagreements` disagreements (Cohen's kappa = `r printp(kappa.rep)`), all were resolved by discussion.

@@

To explain the $`r printnum((prop.support.SR - prop.support.RR)*100)` \%$ gap between standard reports and Registered Reports, we must assume some combination of differences in bias, statistical power, or the proportion of true hypotheses researchers choose to examine. 
Figure\ \@ref(fig:powerbaserate) visualises the combinations of statistical power and proportion of true hypotheses that would produce the observed positive result rates if the literature were completely unbiased. 
For example, if we assume no publication bias and no QRPs, more than $90\%$ of the hypotheses authors of standard reports study must be true and study designs must have more than $90\%$ power for the true effect size. 
This is highly unlikely, meaning that the standard literature is unlikely to reflect reality. 
As we noted above, there is good reason to assume that methodological rigour and statistical power in Registered Reports are as high as in standard reports or higher, leaving the rate of true hypotheses and bias as remaining explanations.

>>>>

Assuming a constant alpha level, the rate of positive results in a literature (i.e., the proportion of supported hypotheses among all tested hypotheses) is influenced by three factors: the proportion of true hypotheses among all tested hypotheses, statistical power, and bias. 
The Registered Reports format combines powerful safeguards against publication bias and QRPs with standards for research quality that are at least equal to ordinary peer review, and often include statistical power requirements that likely exceed those in the standard literature [see e.g., @Maxwell2004; @Szucs2017; @SingletonThorn2019]. 
Therefore, if the emerging Registered Reports literature in Psychology contains fewer positive results than the standard literature, the cause must be either the difference in bias or a lower proportion of true hypotheses tested in Registered Reports (or a combination of the two). 
At this time, we have good reasons to believe in a difference in bias, but less reason to believe in a difference in the proportion of true hypotheses (at least regarding original hypotheses, see below), which would make bias a more plausible explanation of a potential difference in the positive result rate. 
Considering the high standards for research quality in Registered Reports, a large difference in positive results between Registered Reports and the standard literature would also mean that publication bias is not a desirable filter for poorly conducted studies, but that we should worry about the high-quality negative results we are missing because of it.

@@

For standard reports we downloaded a current version of the Essential Science Indicators (ESI) database (retrieved on 4th December 2018) and used Web of Science to search for articles published between 2013 and 2018 with a Boolean search query containing the phrase "test$^\ast$ the hypothes$^\ast$" and the ISSNs of all journals listed in the ESI Psychiatry/Psychology category. 
Using the same sample size as @Fanelli2010, we randomly selected 150 papers from the 1919 search results using the sample() function in R and the seed "20190120" (seed was not preregistered, but no other seeds were tried).
We initially excluded eight papers and replaced them (decision to replace excluded papers was not preregistered) through the same random sampling procedure until 150 studies were found that met our criteria, but later found that two papers had been excluded erroneously, leading to a final sample size of 152 (see Fig.\ \@ref(fig:sampling)).

@@

For Registered Reports we aimed to include all published Registered Reports in the field of Psychology that tested at least one hypothesis, regardless of whether or not they used the phrase "test$^\ast$ the hypothes$^\ast$". 
We downloaded a database of published Registered Reports curated by the Center for Open Science^[https://www.zotero.org/groups/479248/osf/items/collectionKey/KEJP68G9] (retrieved on 19th November 2018), and excluded papers published in journals that were listed in categories other than "Psychiatry/Psychology" or "Multidisciplinary" in the ESI. 
Note that the decision to focus only on the Psychiatry/Psychology category meant excluding 13 Registered Reports published in Cortex because the ESI counts this journal towards the separate category "Neuroscience and Behavior". 
Papers published in multidisciplinary journals and in journals not included in the ESI (e.g., Royal Society Open Science) were hand-coded by AS. 
This deviates from our preregistration insofar as we had not specified how discipline membership would be determined.

@@

### Replication status
We also wanted to code if a study was a replication of a previously published one: 
We expected a much larger proportion of Registered Reports to be direct replications, many of which may have been motivated by scepticism of the original study. 
A lower positive result rate in Registered Reports could then be an effect of failed replications rather than an effect of safeguards against QRPs and publication bias. 
After an initial coding attempt with ill-defined coding criteria had led to too many disagreements (described further in the Appendix), we developed the following strategy (not pre-registered): 
We searched the full texts of all papers for the string "replic$^\ast$" [cf. @Makel2012; @Kohler2019; @Mueller-Langer2019; @Pridemore2018], and for papers that did contain it determined whether the coded hypothesis was a close replication with the goal to verify a previously published result. 
Conceptual replications and internal replications (replication of a study in the same paper) were not counted as replications in this narrow sense, since both are more likely to be motivated by the goal to build on previous work than by scepticism.
AS coded all papers, DL double-coded `r rep.RR.coded.DL` Registered Reports ($`r printnum(rep.RR.coded.DL/sum(included$is_RR)*100)` \%$) and `r rep.SR.coded.DL` standard reports ($`r printnum(rep.SR.coded.DL/sum(included$is_RR==0)*100)` \%$). 
There were `r rep.disagreements` disagreements (Cohen's kappa = `r printp(kappa.rep)`), all were resolved by discussion.

@@

To explain the $`r printnum((prop.support.SR - prop.support.RR)*100)` \%$ gap between standard reports and Registered Reports, we must assume some combination of differences in bias, statistical power, or the proportion of true hypotheses researchers choose to examine. 
Figure\ \@ref(fig:powerbaserate) visualises the combinations of statistical power and proportion of true hypotheses that would produce the observed positive result rates if the literature were completely unbiased. 
For example, assuming no publication bias and no QRPs, even if all hypotheses authors of standard reports tested were true, their study designs would need to have more than $90\%$ power for the true effect size. 
This is highly unlikely, meaning that the standard literature is unlikely to reflect reality. 
As we noted above, there is good reason to assume that methodological rigour and statistical power in Registered Reports are as high as in standard reports or higher, leaving the rate of true hypotheses and bias as remaining explanations.

### 2020-01-14 preprint/manuscript.tex
changed one sentence in the Discussion and two in the description of replication status coding

Assuming a constant alpha level, the rate of positive results in a literature (i.e., the proportion of supported hypotheses among all tested hypotheses) is influenced by three factors: the proportion of true hypotheses among all tested hypotheses, statistical power, and bias.
The Registered Reports format combines powerful safeguards against publication bias and QRPs with standards for research quality that are at least equal to ordinary peer review, and often include statistical power requirements that likely exceed those in the standard literature (see e.g., Maxwell, 2004; Szucs \& Ioannidis, 2017).
Therefore, if the emerging Registered Reports literature in Psychology contains fewer positive results than the standard literature, the cause must be either the difference in bias or a lower proportion of true hypotheses tested in Registered Reports (or a combination of the two).
At this time, we have good reasons to believe in a difference in bias, but less reason to believe in a difference in the proportion of true hypotheses (at least regarding original hypotheses, see below), which would make bias a more plausible explanation of a potential difference in the positive result rate.
Considering the high standards for research quality in Registered Reports, a large difference in positive results between Registered Reports and the standard literature would also mean that publication bias is not a desirable filter for poorly conducted studies, but that we should worry about the high-quality negative results we are missing because of it.

@@

For standard reports we downloaded a current version of the Essential Science Indicators (ESI) database (retrieved on 4th December 2018) and used Web of Science to search for articles published between 2013 and 2018 with a Boolean search query containing the phrase \enquote{test\(^\ast\) the hypothes\(^\ast\)} and the ISSNs of all journals listed in the ESI Psychiatry/Psychology category.
Using the same sample size as Fanelli (2010), we randomly selected 150 papers from the 1919 search results using the sample() function in R and the seed \enquote{20190120} (not specified in our preregistration, but no other seeds were tried).
We initially excluded eight papers and replaced them (the decision to replace excluded papers was not preregistered) through the same random sampling procedure until 150 studies were found that met our criteria, but later found that two papers had erroneously been excluded, leading to a final sample size of 152 rather than 150 (see Fig.~\ref{fig:sampling}).

@@

For Registered Reports we aimed to include all published Registered Reports in the field of Psychology that tested at least one hypothesis, regardless of whether or not they used the phrase \enquote{test\(^\ast\) the hypothes\(^\ast\)}.
We downloaded a database of published Registered Reports curated by the Center for Open Science (\url{https://www.zotero.org/groups/479248/osf/items/collectionKey/KEJP68G9}, retrieved on 19th November 2018), and excluded papers published in journals that were listed in categories other than \enquote{Psychiatry/Psychology} or \enquote{Multidisciplinary} in the ESI.
Note that the decision to focus only on the Psychiatry/Psychology category meant excluding 13 Registered Reports published in Cortex because the ESI counts this journal towards the separate category \enquote{Neuroscience and Behavior}.
Papers published in multidisciplinary journals and in journals not included in the ESI (e.g., Royal Society Open Science) were hand-coded by AS.
This deviates from our preregistration insofar as we had not specified how discipline membership would be determined.

@@

We also wanted to code if a study was a replication of a previously published one:
We expected a much larger proportion of Registered Reports to be direct replications, many of which may have been motivated by scepticism of the original study.
A lower positive result rate in Registered Reports could then be an effect of failed replications rather than an effect of safeguards against QRPs and publication bias.
After an initial coding attempt with ill-defined coding criteria had led to too many disagreements (described further in the Appendix), we developed the following strategy (not pre-registered):
We searched the full texts of all papers for the string \enquote{replic\(^\ast\)} (cf. Makel, Plucker, \& Hegarty, 2012; Köhler \& Cortina, 2019; Mueller-Langer, Fecher, Harhoff, \& Wagner, 2019; Pridemore, Makel, \& Plucker, 2018), and for papers that did contain it determined whether the first hypothesis was a close or direct replication of previously published work.
Conceptual replications and internal replications were not counted as replications in this narrow sense:
The reason to code replication status was the idea that direct replications of previous work may be motivated by replicators' scepticism in the hypothesis and thus have a low prior probability, whereas conceptual replications are usually considered as building on previous work rather than verifying it.
AS coded all papers, DL double-coded 32 Registered Reports (\(45.07 \%\)) and 99 standard reports (\(65.13 \%\)).
There were 5 disagreements (Cohen's kappa = .878), all were resolved by discussion.

@@

To explain the \(52.39 \%\) gap between standard reports and Registered Reports, we must assume some combination of differences in bias, statistical power, or the proportion of true hypotheses researchers choose to examine.
Figure~\ref{fig:powerbaserate} visualises the combinations of statistical power and proportion of true hypotheses that would produce the observed positive result rates if the literature were completely unbiased.
For example, if we assume no publication bias and no QRPs, more than \(90\%\) of the hypotheses authors of standard reports study must be true and study designs must have more than \(90\%\) power for the true effect size.
This is highly unlikely, meaning that the standard literature is unlikely to reflect reality.
As we noted above, there is good reason to assume that methodological rigour and statistical power in Registered Reports are as high as in standard reports or higher, leaving the rate of true hypotheses and bias as remaining explanations.

@@

>>>>

Assuming a constant alpha level, the rate of positive results in a literature (i.e., the proportion of supported hypotheses among all tested hypotheses) is influenced by three factors: the proportion of true hypotheses among all tested hypotheses, statistical power, and bias.
The Registered Reports format combines powerful safeguards against publication bias and QRPs with standards for research quality that are at least equal to ordinary peer review, and often include statistical power requirements that likely exceed those in the standard literature (see e.g., Maxwell, 2004; Singleton Thorn, Dudgeon, \& Fidler, 2019; Szucs \& Ioannidis, 2017).
Therefore, if the emerging Registered Reports literature in Psychology contains fewer positive results than the standard literature, the cause must be either the difference in bias or a lower proportion of true hypotheses tested in Registered Reports (or a combination of the two).
At this time, we have good reasons to believe in a difference in bias, but less reason to believe in a difference in the proportion of true hypotheses (at least regarding original hypotheses, see below), which would make bias a more plausible explanation of a potential difference in the positive result rate.
Considering the high standards for research quality in Registered Reports, a large difference in positive results between Registered Reports and the standard literature would also mean that publication bias is not a desirable filter for poorly conducted studies, but that we should worry about the high-quality negative results we are missing because of it.

@@

For standard reports we downloaded a current version of the Essential Science Indicators (ESI) database (retrieved on 4th December 2018) and used Web of Science to search for articles published between 2013 and 2018 with a Boolean search query containing the phrase \enquote{test\(^\ast\) the hypothes\(^\ast\)} and the ISSNs of all journals listed in the ESI Psychiatry/Psychology category.
Using the same sample size as Fanelli (2010), we randomly selected 150 papers from the 1919 search results using the sample() function in R and the seed \enquote{20190120} (seed was not preregistered, but no other seeds were tried).
We initially excluded eight papers and replaced them (decision to replace excluded papers was not preregistered) through the same random sampling procedure until 150 studies were found that met our criteria, but later found that two papers had been excluded erroneously, leading to a final sample size of 152 (see Fig.~\ref{fig:sampling}).

@@

For Registered Reports we aimed to include all published Registered Reports in the field of Psychology that tested at least one hypothesis, regardless of whether or not they used the phrase \enquote{test\(^\ast\) the hypothes\(^\ast\)}.
We downloaded a database of published Registered Reports curated by the Center for Open Science\footnote{\url{https://www.zotero.org/groups/479248/osf/items/collectionKey/KEJP68G9}} (retrieved on 19th November 2018), and excluded papers published in journals that were listed in categories other than \enquote{Psychiatry/Psychology} or \enquote{Multidisciplinary} in the ESI.
Note that the decision to focus only on the Psychiatry/Psychology category meant excluding 13 Registered Reports published in Cortex because the ESI counts this journal towards the separate category \enquote{Neuroscience and Behavior}.
Papers published in multidisciplinary journals and in journals not included in the ESI (e.g., Royal Society Open Science) were hand-coded by AS.
This deviates from our preregistration insofar as we had not specified how discipline membership would be determined.

@@

We also wanted to code if a study was a replication of a previously published one:
We expected a much larger proportion of Registered Reports to be direct replications, many of which may have been motivated by scepticism of the original study.
A lower positive result rate in Registered Reports could then be an effect of failed replications rather than an effect of safeguards against QRPs and publication bias.
After an initial coding attempt with ill-defined coding criteria had led to too many disagreements (described further in the Appendix), we developed the following strategy (not pre-registered):
We searched the full texts of all papers for the string \enquote{replic\(^\ast\)} (cf. Makel, Plucker, \& Hegarty, 2012; Köhler \& Cortina, 2019; Mueller-Langer, Fecher, Harhoff, \& Wagner, 2019; Pridemore, Makel, \& Plucker, 2018), and for papers that did contain it determined whether the coded hypothesis was a close replication with the goal to verify a previously published result.
Conceptual replications and internal replications (replication of a study in the same paper) were not counted as replications in this narrow sense, since both are more likely to be motivated by the goal to build on previous work than by scepticism.
AS coded all papers, DL double-coded 32 Registered Reports (\(45.07 \%\)) and 99 standard reports (\(65.13 \%\)).
There were 5 disagreements (Cohen's kappa = .878), all were resolved by discussion.

@@

To explain the \(52.39 \%\) gap between standard reports and Registered Reports, we must assume some combination of differences in bias, statistical power, or the proportion of true hypotheses researchers choose to examine.
Figure~\ref{fig:powerbaserate} visualises the combinations of statistical power and proportion of true hypotheses that would produce the observed positive result rates if the literature were completely unbiased.
For example, assuming no publication bias and no QRPs, even if all hypotheses authors of standard reports tested were true, their study designs would need to have more than \(90\%\) power for the true effect size.
This is highly unlikely, meaning that the standard literature is unlikely to reflect reality.
As we noted above, there is good reason to assume that methodological rigour and statistical power in Registered Reports are as high as in standard reports or higher, leaving the rate of true hypotheses and bias as remaining explanations.

@@

\leavevmode\hypertarget{ref-SingletonThorn2019}{}%
Singleton Thorn, F., Dudgeon, P., \& Fidler, F. (2019, September). \emph{Statistical power and effect sizes in psychology are decreasing over time}. Presented at the Metascience 2019 Symposium, Stanford, CA, USA. Retrieved from \url{https://www.metascience2019.org/poster-session/felix-singleton-thorn/}
